---
title: Processing the Romanov brain dataset
author: Aaron Lun
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Downloading the count data

We obtain a single-cell RNA sequencing dataset of the mouse hypothalamus from [Romanov et al. (2017)](https://doi.org/10.1038%2Fnn.4462).
Counts for endogenous genes are available from the Gene Expression Omnibus
using the accession number [GSE74672](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE74672).
We download and cache it using the `r Biocpkg("BiocFileCache")` package.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask = FALSE)
base.url <- "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE74nnn/GSE74672/suppl/"
count.file <- bfcrpath(bfc, paste0(base.url, "GSE74672_expressed_mols_with_classes.xlsx.gz"))
```

We load them into memory.

```{r}
library(R.utils)
tmp.file <- tempfile(fileext=".xlsx")
gunzip(count.file, destname=tmp.file, remove=FALSE)

# WARNING: this requires >20 GB of memory!
library(readxl)
all.data <- read_xlsx(tmp.file, col_names=FALSE)
dim(all.data)
```

# Extracting the metadata

The first 9 rows correspond to metadata and need to be extracted.

```{r}
coldata <- all.data[1:9,]
colfields <- coldata[,1,drop=TRUE]
coldata <- coldata[,-1]

coldata <- t(coldata)
coldata <- data.frame(coldata, stringsAsFactors=FALSE)
rownames(coldata) <- NULL

library(S4Vectors)
coldata <- DataFrame(coldata)
colnames(coldata) <- colfields

for (i in 4:9) {
    coldata[,i] <- as.numeric(coldata[,i])
}
coldata
```

The next three rows are empty, so we just skip them.
We convert the remaining counts into a sparse matrix.

```{r}
raw.counts <- all.data[-(1:12),]
counts <- as.matrix(raw.counts[,-1])

library(Matrix)
counts <- as(counts, "dgCMatrix")
rownames(counts) <- raw.counts[,1,drop=TRUE]
colnames(counts) <- NULL
dim(counts)
```

# Saving to file 

Putting everything together into a `SingleCellExperiment`:

```{r}
library(SingleCellExperiment)
sce <- SingleCellExperiment(list(counts=counts), colData=coldata)

# Splitting off the ERCCs.
status <- ifelse(grepl("^ERCC-[0-9]+", rownames(sce)), "ERCC", "endogenous")
sce <- splitAltExps(sce, status, ref="endogenous")

library(scRNAseq)
sce <- polishDataset(sce)
sce
```

We now save the object to disk in preparation for upload.

```{r}
unlink("finished", recursive=TRUE)
saveDataset(sce, "finished", 
    createMetadata(
        title="Molecular interrogation of hypothalamic organization reveals distinct dopamine neuronal subtypes",
        description="The hypothalamus contains the highest diversity of neurons in the brain. Many of these neurons can co-release neurotransmitters and neuropeptides in a use-dependent manner. Investigators have hitherto relied on candidate protein-based tools to correlate behavioral, endocrine and gender traits with hypothalamic neuron identity. Here we map neuronal identities in the hypothalamus by single-cell RNA sequencing. We distinguished 62 neuronal subtypes producing glutamatergic, dopaminergic or GABAergic markers for synaptic neurotransmission and harboring the ability to engage in task-dependent neurotransmitter switching. We identified dopamine neurons that uniquely coexpress the Onecut3 and Nmur2 genes, and placed these in the periventricular nucleus with many synaptic afferents arising from neuromedin S+ neurons of the suprachiasmatic nucleus. These neuroendocrine dopamine cells may contribute to the dopaminergic inhibition of prolactin secretion diurnally, as their neuromedin S+ inputs originate from neurons expressing Per2 and Per3 and their tyrosine hydroxylase phosphorylation is regulated in a circadian fashion. Overall, our catalog of neuronal subclasses provides new understanding of hypothalamic organization and function.",
        taxonomy.id="10090",
        genome="GRCm38",
        sources=list(
            list(provider="GEO", id="GSE74672"),
            list(provider="PubMed", id="27991900")
        ),
        maintainer.name="Aaron Lun",
        maintainer.email="infinite.monkeys.with.keyboards@gmail.com"
    )
)
```

# Session information

```{r}
sessionInfo()
```
