---
title: Processing the Chen brain dataset
author: Aaron Lun
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Downloading the count data

We obtain a single-cell RNA sequencing dataset of the mouse brain from [Chen et al. (2017)](https://doi.org/10.1016/j.celrep.2017.03.004).
Counts for endogenous genes are available from the Gene Expression Omnibus
using the accession number [GSE87544](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE87544).
We download and cache them using the `r Biocpkg("BiocFileCache")` package.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask = FALSE)
base.url <- paste0("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE87nnn/GSE87544/suppl/"
count.file <- bfcrpath(bfc, paste0(base.url, "GSE87544_Merged_17samples_14437cells_count.txt.gz"))
```

Reading them in as a sparse matrix.

```{r}
library(scuttle)
counts <- readSparseCounts(count.file)
dim(counts)
```

# Downloading the metadata

We also download the cluster identities.

```{r}
cluster.file <- bfcrpath(bfc, paste0(base.url, "GSE87544_1443737Cells.SVM.cluster.identity.renamed.csv.gz"))
coldata <- read.csv(cluster.file, stringsAsFactors=FALSE, row.names=1)
coldata <- as(coldata, "DataFrame")
coldata
```

We check that the columns are in the same order.

```{r}
m <- match(colnames(counts), rownames(coldata))
coldata <- coldata[m,]
stopifnot(identical(colnames(counts), rownames(coldata)))
```

# Saving to file 

We slap together a `SingleCellExperiment` object:

```{r}
library(scRNAseq)
sce <- SingleCellExperiment(list(counts=counts), colData=coldata)
sce <- polishDataset(sce)
sce
```

Now, saving to disk:

```{r}
saveDataset(sce, "finished",
    createMetadata(
        title="Single-Cell RNA-Seq Reveals Hypothalamic Cell Diversity",
        description="The hypothalamus is one of the most complex brain structures involved in homeostatic regulation. Defining cell composition and identifying cell-type-specific transcriptional features of the hypothalamus is essential for understanding its functions and related disorders. Here, we report single-cell RNA sequencing results of adult mouse hypothalamus, which defines 11 non-neuronal and 34 neuronal cell clusters with distinct transcriptional signatures. Analyses of cell-type-specific transcriptomes reveal gene expression dynamics underlying oligodendrocyte differentiation and tanycyte subtypes. Additionally, data analysis provides a comprehensive view of neuropeptide expression across hypothalamic neuronal subtypes and uncover Crabp1+ and Pax6+ neuronal populations in specific hypothalamic sub-regions. Furthermore, we found food deprivation exhibited differential transcriptional effects among the different neuronal subtypes, suggesting functional specification of various neuronal subtypes. Thus, the work provides a comprehensive transcriptional perspective of adult hypothalamus, which serves as a valuable resource for dissecting cell-type-specific functions of this complex brain region.",
        taxonomy.id="10090",
        genome="GRCm38",
        sources=list(
            list(provider="GEO", id='GSE87544'),
            list(provider="PubMed", id='28355573')
        ),
        maintainer.name="Aaron Lun",
        maintainer.email="infinite.monkeys.with.keyboards@gmail.com"
    )
)
```

# Session information

```{r}
sessionInfo()
```
