---
title: Processing the Hu cortex snRNA-seq data
author: "Aaron Lun"
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Download the data

We obtain a single-nucleus RNA sequencing dataset of mouse brains from [Hu et al. (2017)](https://doi.org/10.1016/j.molcel.2017.11.017).
Counts for endogenous genes and antibody-derived tags are available from the Gene Expression Omnibus
using the accession number [GSE106678](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE106678).

```{r}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask = FALSE)
tarred <- bfcrpath(bfc, "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE106678&format=file")

temp <- tempfile()
dir.create(temp)
untar(tarred, exdir=temp)
list.files(temp)
```

We also create a metadata template:

```{r}
meta <- list(
    title="Dissecting cell-type composition and activity-dependent transcriptional state in mammalian brains by massively parallel single-nucleus RNA-Seq [%s only]",
    description="Massively parallel single-cell RNA sequencing can precisely resolve cellular diversity in a high-throughput manner at low cost, but unbiased isolation of intact single cells from complex tissues, such as adult mammalian brains, is challenging. Here, we integrate sucrose-gradient assisted purification of nuclei with droplet microfluidics to develop a highly scalable single-nucleus RNA-Seq approach (sNucDrop-Seq), which is free of enzymatic dissociation and nuclei sorting. By profiling ~18,000 nuclei isolated from cortical tissues of adult mice, we demonstrate that sNucDrop-Seq not only accurately reveals neuronal and non-neuronal subtype composition with high sensitivity, but also enables in-depth analysis of transient transcriptional states driven by neuronal activity, at single-cell resolution, in vivo.

Maintainer note: this dataset contains only the %s cells from the study.",
    taxonomy_id="10090",
    genome="GRCm38",
    sources=list(
        list(provider="GEO", id="GSE106678"),
        list(provider="PubMed", id="29220646")
    ),
    maintainer_name="Aaron Lun",
    maintainer_email="infinite.monkeys.with.keyboards@gmail.com"
)
```

We clear out any existing output directory:

```{r}
unlink("finished", recursive=TRUE)
dir.create("finished")
```

# Processing the data

We load in each of the files.

```{r}
library(scuttle)
counts <- list()
for (x in list.files(temp, full.names=TRUE)) {
    prefix <- sub("^[^_]*_", "", x)
    prefix <- sub("_.*", "", prefix)
    counts[[prefix]] <- readSparseCounts(x)
}
do.call(rbind, lapply(counts, dim))
```

For some unknown reason, each matrix has its own set of features!
Crazy.
At least the intersection is of a reasonable size.

```{r}
length(Reduce(intersect, lapply(counts, rownames)))
```

I can't be sure that the missing features in a given matrix have all-zero expression values.
So, I will make the executive decision of taking the intersection across matrices and `cbind`'ing them together into a single matrix for each system (i.e., cortex or 3T3).
If you don't like that, take it up with the authors.

# Saving the 3T3 data

Combining the 3T3 matrices for cell and nuclei:

```{r}
counts.3T3 <- counts[c("cell-3T3", "nuclei-3T3")]
common.3T3 <- Reduce(intersect, lapply(counts.3T3, rownames))
combined.3T3 <- do.call(cbind, lapply(counts.3T3, function(x) x[common.3T3,]))
dim(combined.3T3)
```

Slapping together a `SingleCellExperiment`:

```{r}
library(SingleCellExperiment)
sce <- SingleCellExperiment(list(counts=combined.3T3))
sce$protocol <- rep(sub("-3T3$", "", names(counts.3T3)), vapply(counts.3T3, ncol, 0L))
table(sce$protocol)
```

Adding some polisht o optimize for space:

```{r}
library(scRNAseq)
sce <- polishDataset(sce)
sce
```

And saving it:

```{r}
copy <- meta
copy$title <- sprintf(copy$title, "3T3")
copy$description <- sprintf(copy$description, "3T3")
saveDataset(sce, "finished/3T3", copy)
```

# Saving the cortex data

Defining the sample identities:

```{r}
samples <- c(
    "nuclei-ctx-1",
    "nuclei-ctx-2",
    "nuclei-ctx-3", 
    "nuclei-ctx-4",
    "nuclei-ctx-5",
    "nuclei-ctx-6",
    "nuclei-ctx-7",
    "nuclei-ctx-8",
    "nuclei-ctx-9",
    "nuclei-ctx-10",
    "nuclei-ctx-11",
    "nuclei-ctx-12",
    "nuclei-ctx-13",
    "nuclei-ctx-saline1", 
    "nuclei-ctx-PTZ1", 
    "nuclei-ctx-saline2",
    "nuclei-ctx-PTZ2" 
)

# Checking that everyone is accounted for:
cortex.samples <- sort(names(counts)[grep("^nuclei-ctx-", names(counts))])
stopifnot(identical(cortex.samples, sort(samples)))
```

Combining the cortex matrices across samples:

```{r}
counts.cortex <- counts[samples]
common.cortex <- Reduce(intersect, lapply(counts.cortex, rownames))
combined.cortex <- do.call(cbind, lapply(counts.cortex, function(x) x[common.cortex,]))
dim(combined.cortex)
```

Slapping together a `SingleCellExperiment`:

```{r}
library(SingleCellExperiment)
sce <- SingleCellExperiment(list(counts=combined.cortex))
```

Adding some more information about each sample.

```{r}
replicator <- rep(seq_along(samples), vapply(counts.cortex, ncol, 0L))

treatment <- rep(NA_character_, length(samples))
treatment[grep("^nuclei-ctx-saline[0-9]+$", samples)] <- "saline"
treatment[grep("^nuclei-ctx-PTZ[0-9]+$", samples)] <- "PTZ"
sce$treatment <- treatment[replicator]
table(sce$treatment, useNA="always")

animal <- as.integer(sub("[^0-9]+([0-9]+)$", "\\1", samples))
stopifnot(!anyNA(animal))
sce$animal <- animal[replicator]
table(sce$animal)
```

Polishing it up:

```{r}
sce <- polishDataset(sce)
sce
```

And saving it:

```{r}
copy <- meta
copy$title <- sprintf(copy$title, "cortex")
copy$description <- sprintf(copy$description, "cortex")
saveDataset(sce, "finished/cortex", copy)
```

# Session information {-}

```{r}
sessionInfo()
```
