---
title: Fixing the Stoeckius hashing dataset
author: Aaron Lun
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Promoting to `SingleCellExperiment`

For some reason, I saved all of the to-be-alternative experiments as `SummarizedExperiment`s instead of `SingleCellExperiment`s.
I was probably hoping to eliminate the potential for unnecessary recursion of alternative experiments, but in doing so,
got rid of useful features for `SingleCellExperiments` like reduced dimensions and size factors.

So let's put that stuff back.
First we clone the project:

```{r}
tmp <- "update_20240418"
unlink(tmp, recursive=TRUE) # get rid of any previous directory.
gypsum::cloneVersion("scRNAseq", "stoeckius-hashing-2018", "2023-12-20", destination=tmp)
```

Now we go in and promote the `OBJECT` file to a `SingleCellExperiment`, 
following the specification [here](https://github.com/ArtifactDB/takane/blob/gh-pages/docs/specifications/single_cell_experiment/1.0.md).
There's no need to do anything else as the new SCE just differs from the old SE by the class definition, not any of its contents.

```{r}
update <- function(obj.path) {
    obj <- jsonlite::fromJSON(obj.path, simplifyVector=FALSE)
    obj$type <- "single_cell_experiment"
    obj$ranged_summarized_experiment <- list(version="1.0")
    obj$single_cell_experiment <- list(version="1.0")

    # Replacing the link with the new content; it's important to delete the link
    # before doing so, otherwise we would accidentally modify the cache!
    unlink(obj.path) 
    write(jsonlite::toJSON(obj, auto_unbox=TRUE, pretty=4), file=obj.path)
}
```

Running our update script on all of the alternative experiments:

```{r}
for (x in c("pbmc/adt", "pbmc/igg", "pbmc/hto", "mixture/hto")) {
    update(file.path(tmp, x, "OBJECT"))
}
```

We validate and load each one, just to check that it's all good.

```{r}
library(alabaster.base)
validateObject(file.path(tmp, "pbmc/adt"))
out <- readObject(file.path(tmp, "pbmc/adt"))
stopifnot(is(out, "SingleCellExperiment"))
out

validateObject(file.path(tmp, "pbmc/igg"))
out <- readObject(file.path(tmp, "pbmc/igg"))
stopifnot(is(out, "SingleCellExperiment"))
out

validateObject(file.path(tmp, "pbmc/hto"))
out <- readObject(file.path(tmp, "pbmc/hto"))
stopifnot(is(out, "SingleCellExperiment"))
out

validateObject(file.path(tmp, "mixture/hto"))
out <- readObject(file.path(tmp, "mixture/hto"))
stopifnot(is(out, "SingleCellExperiment"))
out
```

# Session information {-}

```{r}
sessionInfo()
```
