---
title: Processing the Tasic brain dataset
author: Aaron Lun
date: "June 8, 2019"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
bibliography: ref.bib
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Overview

The @tasic2016adult study provides a single-cell RNA seq dataset involving the mouse brain. 
This contains approximately 1800 cells of varying types such as oligodendrocytes, microglia and neurons.
Individual cells were isolated by FACS and library preparation was performed on each cell using the SMARTer kit.
After sequencing, expression was quantified by counting the number of reads mapped to each gene.

# Downloading the count data

Counts for endogenous genes and spike-in transcripts are available from the Gene Expression Omnibus
using the accession number [GSE71585](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE71585).
We download and cache them using the `r Biocpkg("BiocFileCache")` package.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask = FALSE)
base.url <- file.path("ftp://ftp.ncbi.nlm.nih.gov/geo/series",
    "GSE71nnn/GSE71585/suppl")
count.file <- bfcrpath(bfc, file.path(base.url,
    "GSE71585_RefSeq_counts.csv.gz"))
spike.file <- bfcrpath(bfc, file.path(base.url,
    "GSE71585_ERCC_and_tdTomato_counts.csv.gz"))
```

We load them into memory.

```{r}
count.mat <- read.csv(count.file, row.names=1, header=TRUE, check.names=FALSE)
count.mat <- as.matrix(count.mat)
dim(count.mat)
spike.mat <- read.csv(spike.file, row.names=1, header=TRUE, check.names=FALSE)
spike.mat <- as.matrix(spike.mat)
dim(spike.mat)
```

We check that all objects are in the same order and combine them.

```{r}
stopifnot(identical(colnames(count.mat), colnames(spike.mat)))
counts <- rbind(count.mat, spike.mat)
```

# Downloading the per-cell metadata

We also download a file containing the metadata for each cell.

```{r}
meta.file <- bfcrpath(bfc, file.path(base.url,
    "GSE71585_Clustering_Results.csv.gz"))
metadata <- read.csv(meta.file, stringsAsFactors=FALSE)
nrow(metadata)
head(metadata)
```

Some clean-up is necessary to replace `"N/A"` with actual `NA_character_` entries, 
which are more appropriate for conveying missingness.

```{r}
for (i in colnames(metadata)) {
    current <- metadata[,i]
    to.rename <- current %in% c("N/A")
    current[to.rename] <- NA
    metadata[,i] <- current
}
```

We check that all objects are in the same order, and use this to create a column-level `DataFrame`.

```{r}
m <- match(colnames(counts), metadata$sample_title)
stopifnot(all(!is.na(m)))
metadata <- metadata[m,]
library(S4Vectors)
coldata <- as(metadata, "DataFrame")
```

# Saving to file 

We now save all of the components to file for upload to `r Biocpkg("ExperimentHub")`.
These will be used to construct a `SingleCellExperiment` on the client side when the dataset is requested.

```{r}
path <- file.path("scRNAseq", "tasic-brain", "2.0.0")
dir.create(path, showWarnings=FALSE, recursive=TRUE)
saveRDS(counts, file=file.path(path, "counts.rds"))
saveRDS(coldata, file=file.path(path, "coldata.rds"))
```

# Session information

```{r}
sessionInfo()
```

# References
