---
title: Processing the Macosko retina dataset
author: Aaron Lun
date: "June 8, 2019"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
bibliography: ref.bib
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Downloading the count data

We obtain a single-cell RNA sequencing dataset of the mouse retina from @shekhar2016comprehensive. 
Counts for endogenous genes and spike-in transcripts are available from the Gene Expression Omnibus
using the accession number [GSE81904](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE81904).
We download and cache them using the `r Biocpkg("BiocFileCache")` package.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask = FALSE)
count.url <- file.path("ftp://ftp.ncbi.nlm.nih.gov/geo/series",
    "GSE81nnn/GSE81904/suppl/GSE81904_BipolarUMICounts_Cell2016.txt.gz")
count.file <- bfcrpath(bfc, count.url)
```

We load them into memory.

```{r}
library(scater)
counts <- readSparseCounts(count.file)
dim(counts)
```

# Downloading the per-cell metadata

We also download a file containing the metadata for each cell.
(Courtesy of Vladimir Kiselev and Martin Hemberg, as the original annotation has disappeared.)

```{r}
meta.file <- bfcrpath(bfc, file.path("https://s3.amazonaws.com",
    "scrnaseq-public-datasets/manual-data/shekhar/clust_retinal_bipolar.txt"))
coldata <- read.delim(meta.file, stringsAsFactors=FALSE, check.names=FALSE)

library(S4Vectors)
coldata <- as(coldata, "DataFrame")
coldata
```

We match the metadata to the columns.

```{r}
m <- match(colnames(counts), coldata$NAME)
coldata <- coldata[m,]
coldata$NAME <- colnames(counts)
summary(is.na(m))
```

# Saving to file 

We now save all of the components to file for upload to `r Biocpkg("ExperimentHub")`.

```{r}
path <- file.path("scRNAseq", "shekhar-retina", "2.0.0")
dir.create(path, showWarnings=FALSE, recursive=TRUE)
saveRDS(counts, file=file.path(path, "counts.rds"))
saveRDS(coldata, file=file.path(path, "coldata.rds"))
```

# Session information

```{r}
sessionInfo()
```

# References
