---
title: Processing the Lawlor pancreas dataset
author: Aaron Lun
date: "June 27, 2019"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
bibliography: ref.bib
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Downloading the count data

We obtain a single-cell RNA sequencing dataset of human pancreas from @lawlor2017singlecell.
Counts for endogenous genes are available from the Gene Expression Omnibus
using the accession number [GSE86469](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE86469).
We download and cache it using the `r Biocpkg("BiocFileCache")` package.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)    
count.tab <- bfcrpath(bfc, file.path(
    "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE86nnn/GSE86469/suppl",
    "GSE86469_GEO.islet.single.cell.processed.data.RSEM.raw.expected.counts.csv.gz"
))
```

We then load it in as a sparse matrix.

```{r}
library(scater)
counts <- readSparseCounts(count.tab, sep=",", quote='"', row.names=1)
dim(counts)
```

# Downloading the metadata

We extract the metadata for this study using the `r Biocpkg("GEOquery")` package.

```{r}
library(GEOquery)
coldata <- pData(getGEO("GSE86469")[[1]])

library(S4Vectors)
coldata <- as(coldata, "DataFrame")
rownames(coldata) <- NULL
colnames(coldata)
```

We remove the constant columns, as these are unlikely to be interesting.

```{r}
nlevels <- vapply(coldata, FUN=function(x) length(unique(x[!is.na(x)])), 1L)
coldata <- coldata[,nlevels > 1L]
```

We also remove the columns related to the GEO accession itself, as well as some redundant fields.

```{r}
coldata <- coldata[,! colnames(coldata) %in%
    c("geo_accession", "relation", "relation.1")]
coldata <- coldata[,!grepl("^characteristics_ch", colnames(coldata))]
```

We convert all factors into character vectors, and we clean up the column names.

```{r}
for (i in colnames(coldata)) {
    if (is.factor(coldata[[i]])) {
        coldata[[i]] <- as.character(coldata[[i]])
    }
}

stopifnot(identical(colnames(counts), coldata[,1]))
colnames(coldata) <- sub("[:_]ch1$", "", colnames(coldata))
coldata
```

# Saving to file 

We now save all of the components to file for upload to `r Biocpkg("ExperimentHub")`.
These will be used to construct a `SingleCellExperiment` on the client side when the dataset is requested.

```{r}
path <- file.path("scRNAseq", "lawlor-pancreas", "2.0.0")
dir.create(path, showWarnings=FALSE, recursive=TRUE)
saveRDS(counts, file=file.path(path, "counts.rds"))
saveRDS(coldata, file=file.path(path, "coldata.rds"))
```

# Session information

```{r}
sessionInfo()
```

# References
