---
title: Processing the Muraro pancreas dataset
author: Aaron Lun
date: 28 June 2019
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
bibliography: ref.bib
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Downloading the count data

We obtain a single-cell RNA sequencing dataset of human pancreas from @muraro2016singlecell.
A count matrix is provided from the Gene Expression Omnibus under the accession code GSE85241.
We download it using `r Biocpkg("BiocFileCache")` to cache the results:

```{r}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask=FALSE)
muraro.fname <- bfcrpath(bfc, file.path("ftp://ftp.ncbi.nlm.nih.gov/geo/series",
    "GSE85nnn/GSE85241/suppl",
    "GSE85241%5Fcellsystems%5Fdataset%5F4donors%5Fupdated%2Ecsv%2Egz"))
```

We first read the table into memory as a sparse matrix.

```{r}
library(scater)
mat <- readSparseCounts(muraro.fname, quote="\"")
dim(mat)
```

# Loading the column metadata

We extract the metadata from the column names.

```{r}
donor.names <- sub("^(D[0-9]+).*", "\\1", colnames(mat))
plate.id <- sub("^D[0-9]+-([0-9]+)_.*", "\\1", colnames(mat))
table(donor.names, plate.id)
```

We also load in some cell annotations provided by the authors (via Vladimir Wikislev and Martin Hemberg).

```{r}
muraro.cell <- bfcrpath(bfc, file.path("https://s3.amazonaws.com",
    "scrnaseq-public-datasets/manual-data/muraro",
    "cell_type_annotation_Cels2016.csv"))
coldata <- read.delim(muraro.cell, row.names=1, stringsAsFactors=FALSE)

library(S4Vectors)
coldata <- as(coldata, "DataFrame")
coldata
```

We clean up the names of `coldata` and make sure to synchronize the rows with `mat`.

```{r}
colnames(coldata) <- "label"
rownames(coldata) <- sub("\\.", "-", rownames(coldata))
m <- match(colnames(mat), rownames(coldata))
coldata <- coldata[m,,drop=FALSE]
rownames(coldata) <- colnames(mat)
```

Finally, we add the extra annotations.

```{r}
coldata$donor <- donor.names
coldata$plate <- factor(plate.id)
coldata
```

# Processing the row metadata

As an extra service, we will clean up the row metadata as well.

```{r}
symbol <- sub("__.*", "", rownames(mat))
loc <- sub(".*__", "", rownames(mat))
rowdata <- DataFrame(symbol=symbol, chr=loc)
rowdata
```

# Saving to file

We save all of the components to file for upload to `r Biocpkg("ExperimentHub")`.

```{r}
path <- file.path("scRNAseq", "muraro-pancreas", "2.0.0")
dir.create(path, showWarnings=FALSE, recursive=TRUE)
saveRDS(counts, file=file.path(path, "counts.rds"))
saveRDS(rowdata, file=file.path(path, "rowdata.rds"))
saveRDS(coldata, file=file.path(path, "coldata.rds"))
```

# Session information

```{r}
sessionInfo()
```

# References
